---
title: 'Causal generalization in extinction learning: network analysis'
output: html_document
---

```{r setup, include=FALSE}
#knitr options
knitr::opts_chunk$set(echo=TRUE, error=TRUE, warning=FALSE, message=FALSE, fig.align="center")

#load packagesc nm
packages <- c("dplyr", "summarytools", 
              "huge", "qgraph", "bootnet")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))
}
lapply(packages, require, character.only=TRUE)

```

```{r qgraph_setup, include=FALSE}
# setup colorbrewer2.0 palette:
# https://colorbrewer2.org/?type=qualitative&scheme=Set2&n=6
colours_all.set2=c('#66c2a5','#8da0cb','#e78ac3','#fc8d62','#a6d854','#ffd92f')

#max edge weight for constancy of visualisation across networks:
max_edge=0.4
```

```{r load_data, echo=FALSE}
#load data
preTx<-read.csv("../results/pre-tx_master.csv")
```

***

## 1. Psychological symptom data only 

First, lets's select the data we want to include in the network model (for now, only self-reported psychological symptom data).  

```{r echo=FALSE, include=FALSE}
#select desired variables to model:
data<-dplyr::select(preTx, 
                    bl_PCL5_reExp, bl_PCL5_avoid, bl_PCL5_negAff, bl_PCL5_anhed, bl_PCL5_extBeh, 
                    bl_PCL5_anxArou, bl_PCL5_dysArou,
                    bl_BDI_II_cog, bl_BDI_II_physAff
                    )
#view distribution of variables
print(dfSummary(data, max.distinct.values=5, graph.magnif = 0.75), method = 'render')

#probably want to transform the symptom data before network analysis as GGMs assume roughly gaussian distributions: we can do this directly using qgraph npn option
```

Now, let's construct ta Gaussian Graphical Model (GGM) from the data:  

```{r echo=FALSE}
#estimate the network via a Gaussian Graphical Model:
#as above, we will choose a unthresholded network with Pearson correlations
n1<-estimateNetwork(data, 
                    default="EBICglasso",
                    corArgs = list(method = "pearson"), #default as of bootnet1.4
                    corMethod="npn",
                    threshold = FALSE
                    )

#make a pretty version of the GGM network plot:
names<-c('PCL1', 'PCL2', 'PCL3', 'PCL4', 'PCL5', 'PCL6', 'PCL7', 
         'BDI1', 'BDI2'
         )
longnames<-c('re-experiencing', 'avoidance', 'negative affect', 'anhedonia', 
             'externalizing behaviour', 'anxious arousal', 'dysphoric arousal',
             'cognitive symptoms', 'physical/affective symptoms'
             )
gr<-list('PTSD symptoms'=c(1:7),
         'Depression symptoms'=c(8:9)
         )

p1<-plot(n1, layout="spring",
         labels=names, groups=gr, nodeNames=longnames,
         maximum=max_edge,
         vsize=6, cut=0, border.width=1.5, border.color='black', 
         colors=colours_all.set2, #palette="pastel",
         legend.cex=.4, GLratio=2.5, mar=c(2,2,2,2),
         title="")

sum1<-sum(abs(n1$graph[upper.tri(n1$graph, diag=FALSE)])); sum1     #4.10;
```

And investigate stability of the network:
```{r echo=FALSE}
#bootstrap edges
b1n1<-bootnet(n1, nCores=8, nBoots=2000, statistics = c("edge", "strength"))
#save output *immediately* for big datasets as the above step is intensive
save(b1n1, file = "b1n1.Rdata")

#multiverse plot
multiverse(b1n1)

#plot edge weights CI
plot(b1n1, labels = FALSE, order = "sample")

```

***

## 2. Psychological symptoms and extinction task data

```{r echo=FALSE, include=FALSE}
#select desired variables to model:
data<-dplyr::select(preTx, 
                    bl_PCL5_reExp, bl_PCL5_avoid, bl_PCL5_negAff, bl_PCL5_anhed, bl_PCL5_extBeh, 
                    bl_PCL5_anxArou, bl_PCL5_dysArou,
                    bl_BDI_II_cog, bl_BDI_II_physAff,
                    logBF, CSplus_b9, CSminus_b9
                    )

#view distribution of variables
print(dfSummary(data, max.distinct.values=5, graph.magnif = 0.75), method = 'render')
# #probably want to transform the symptom data before network analysis as GGMs assume roughly gaussian distributions
```

Now, let's see what happens when we include extinction task performance variables in the psychological symptom network:  
```{r echo=FALSE}
#as above, we will choose a thresholded network with pearson correlations (the default)
n2<-estimateNetwork(data, 
                    default="EBICglasso",
                    corArgs = list(method = "pearson"),
                    corMethod = "npn",
                    threshold = FALSE
                    )
n2.summ<-n2$graph

#make a pretty version of the GGM network plot:
names<-c('PCL1', 'PCL2', 'PCL3', 'PCL4', 'PCL5', 'PCL6', 'PCL7', 
         'BDI1', 'BDI2', 
         'logBF', 'ext', 'safe'
         )
longnames<-c('re-experiencing', 'avoidance', 'negative affect', 'anhedonia', 
             'externalizing behaviour', 'anxious arousal', 'dysphoric arousal',
             'cognitive symptoms', 'physical/affective symptoms',
             'logBF(alpha>0)', 'extinction resistance', 'safety learning failure'
             )
gr<-list('PTSD symptoms'=c(1:7),
         'Depression symptoms'=c(8:9),
         'Extinction task'=c(10:12)
         )
p2<-plot(n2, layout="spring",
         labels=names, groups=gr, nodeNames=longnames,
         maximum=max_edge,
         vsize=6, cut=0, border.width=1.5, border.color='black', 
         colors=colours_all.set2, #palette="pastel",
         legend.cex=.4, GLratio=2.5, mar=c(2,2,2,2),
         title="")

#get sum of network edges:
sum2<-sum(abs(n2$graph[upper.tri(n2$graph, diag=FALSE)])); sum2    # 5.02;
```

And investigate network stability:

```{r echo=FALSE}
#bootstrap
b1n2<-bootnet(n2, nCores=8, nBoots=2000, statistics = c("edge", "strength"))
#save output *immediately*
save(b1n2, file = "b1n2.Rdata")

#multiverse plot 
multiverse(b1n2)

#plot edge weights CI
plot(b1n2, labels=FALSE, order="sample")
t2<-summary(b1n2) #to generate tibble with output stats; see ?summary.bootnet for details
```

And power:

```{r echo=FALSE}
#simulate 1000 repititions in 8 cores under different sampling levels:
n2.sim<-netSimulator(n2,
                     default = "EBICglasso",
                     nCases = c(56,100,150, 250,500),
                     nReps = 1000,
                     nCores = 8)

#table of results:
n2.sim.res<-n2.sim %>%
            dplyr::group_by(nCases) %>%
            dplyr::summarise(med_corr = median(correlation), iqr_corr = IQR(correlation),
                             med_sens = median(sensitivity), iqr_sens = IQR(sensitivity), 
                             med_spec = median(specificity), iqr_spec = IQR(specificity))
print.data.frame(n2.sim.res)

#plot results:
plot(n2.sim)
```

***

## 3. Psychological symptoms and extinction task data, with demographic data  

```{r echo=FALSE, include=FALSE}
#select desired variables to model:
data<-dplyr::select(preTx, 
                    bl_PCL5_reExp, bl_PCL5_avoid, bl_PCL5_negAff, bl_PCL5_anhed, bl_PCL5_extBeh, 
                    bl_PCL5_anxArou, bl_PCL5_dysArou,
                    bl_BDI_II_cog, bl_BDI_II_physAff,
                    logBF, CSplus_b9, CSminus_b9,
                    pt_age, education,
                    bl_MOS_SS_total, 
                    trauma_total
                    )

#view distribution of variables
print(dfSummary(data, max.distinct.values=5, graph.magnif = 0.75), method = 'render')
# #probably want to transform the symptom data before network analysis as GGMs assume roughly gaussian distributions
```

Now, let's see what happens when we throw everything in:  

```{r echo=FALSE}
#estimate the network via a Gaussian Graphical Model:
#as above, we will choose a unthresholded network with pearson correlations (the default)
n3<-estimateNetwork(data, 
                    default="EBICglasso",
                    corArgs = list(method = "pearson"),
                    corMethod = "npn",
                    threshold = FALSE
                    )
n3.summ<-n3$graph

#make a pretty version of the GGM network plot:
names<-c('PCL1', 'PCL2', 'PCL3', 'PCL4', 'PCL5', 'PCL6', 'PCL7', 
         'BDI1', 'BDI2',
         'logBF', 'ext', 'safe',
         'age', 'edu', 
         'soc',
         'TH'
         )
longnames<-c('re-experiencing', 'avoidance', 'negative affect', 'anhedonia', 
             'externalizing behaviour', 'anxious arousal', 'dysphoric arousal',
             'cognitive symptoms', 'physical/affective symptoms',
             'logBF(alpha>0)', 'extinction resistance', 'safety learning failure',
             'age', 'education level', 
             'perceived social support',
             'additional (nonWTC) trauma'
             )
gr<-list('PTSD symptoms'=c(1:7),
         'Depression symptoms'=c(8:9),
         'Extinction task'=c(10:12),
         'Clinical covariates'=c(13:16)
          )
p3<-plot(n3, layout="spring",
         labels=names, groups=gr, nodeNames=longnames,
         maximum=max_edge,
         vsize=6, cut=0, border.width=1.5, border.color='black', 
         colors=colours_all.set2, #palette="pastel",
         legend.cex=.4, GLratio=2.5, mar=c(2,2,2,2),
         title="")
```

And investigate network stability:

```{r echo=FALSE}
#bootstrap routines, non-parametric bootstrap when handling ordinal data
b1n3<-bootnet(n3, nCores=8, nBoots=2000, statistics = c("edge", "strength"))
#save output *immediately* for big datasets as the above step is intensive
save(b1n3, file = "b1n3.Rdata")

#multiverse plot of edge weight stability
multiverse(b1n3)

#plot edge weights CI
plot(b1n3, labels=FALSE, order="sample")
t3<-summary(b1n3) #to generate tibble with output stats; see ?summary.bootnet for details
```

Investigate change from network 2 (symptom + task data) to network 3 (symptom + task data + covariates), following approach from Armour et al. (2017):

```{r echo=FALSE}
#delete extinction task variables from network of symptom and extinction task variables (cols 10:12)
n3graph.symptoms<-n3$graph[-c(10:16),-c(10:16)] 

#calculate difference between symptoms network in n1 and that in n2
n1n3.delta<-n1$graph - n3graph.symptoms
max(n1n3.delta) #0.10 for BDIcog<->BDIphys
mean(abs(n1n3.delta[upper.tri(n1n3.delta, diag=FALSE)])) #0.061; #delta re-exp:avoid=0.019

#plot difference graph
names<-c('PCL1', 'PCL2', 'PCL3', 'PCL4', 'PCL5', 'PCL6', 'PCL7', 
         'BDI1', 'BDI2')
longnames<-c('re-experiencing', 'avoidance', 'negative affect', 'anhedonia', 
             'externalizing behaviour', 'anxious arousal', 'dysphoric arousal',
             'cognitive symptoms', 'physical/affective symptoms')
gr<-list('PTSD symptoms'=c(1:7),
         'Depression symptoms'=c(8:9))
qgraph(n1n3.delta, layout=p1$layout,
       labels=names, groups=gr, nodeNames=longnames,
       maximum=max_edge,
       vsize=6, cut=0, border.width=1.5, border.color='black', 
       theme="colorblind", colors=colours_all.set2,
       legend.cex=.4, GLratio=2.5, mar=c(2,2,2,2),
       title="")

#correlation between symptoms network in n1 and that in n2
cor(as.vector(n1$graph), as.vector(n3graph.symptoms), method="pearson") #0.987

#mean edge weight among symptoms=0.11; adding ext task measures reduces sum of edges among symptoms by 11.2%
sum1<-sum(abs(n1$graph[upper.tri(n1$graph, diag=FALSE)])); sum1                      #4.10; sum/2 of ptsd network matrix
mean1<-mean(abs(n1$graph[upper.tri(n1$graph,diag=FALSE)])); mean1                    #0.11; mean edge strength
sum2<-sum(abs(n3graph.symptoms[upper.tri(n3graph.symptoms, diag=FALSE)])); sum2      #3.64; sum/2 of ptsd + extinction task matrix, with covariate cells deleted (ptsd network accounting for ext task but without them)
mean2<-mean(abs(n3graph.symptoms[upper.tri(n3graph.symptoms, diag=FALSE)])); mean2   #0.10; mean edge strength
((sum2-sum1)/sum1)*100 #change in connectivity of PTSD network once ext task metrics are added: -11.2%

```

Investigate power:

```{r echo=FALSE}
#simulate 1000 repititions in 8 cores under different sampling levels:
n3.sim<-netSimulator(n3,
                     default = "EBICglasso",
                     nCases = c(56,100,150,250,500),
                     nReps = 1000,
                     nCores = 8)

#table of results:
n3.sim.res<-n3.sim %>%
            dplyr::group_by(nCases) %>%
            dplyr::summarise(med_corr = median(correlation), iqr_corr = IQR(correlation),
                             med_sens = median(sensitivity), iqr_sens = IQR(sensitivity), 
                             med_spec = median(specificity), iqr_spec = IQR(specificity))
print.data.frame(n3.sim.res)

#plot results:
plot(n3.sim)
```

Save R package info for reproducibility:

```{r echo=FALSE}
writeLines(capture.output(sessionInfo()), "RsessionInfo.txt")
```

# References:  

Tutorial on network estimation: Epskamp, S., & Fried, E. I. (2018). A tutorial on regularized partial correlation networks. Psychological Methods, 23(4), 617–634. https://doi.org/10.1037/met0000167 
(see also: Network analysis workshop (FLAMES, Ghent, 2019), by Eiko Fried and Julian Burger https://osf.io/p527f/)

Power analysis using bootnet function netSimulator: Faelens, L., Hoorelbeke, K., Fried, E., De Raedt, R., & Koster, E. H. W. (2019). Negative influences of Facebook use through the lens of network analysis. Computers in Human Behavior, 96, 13–22. https://doi.org/10.1016/j.chb.2019.02.002

Network differences with/without covariates: Armour, C., Fried, E. I., Deserno, M. K., Tsai, J., & Pietrzak, R. H. (2017). A network analysis of DSM-5 posttraumatic stress disorder symptoms and correlates in U.S. military veterans. Journal of Anxiety Disorders, 45, 49–59. https://doi.org/10.1016/j.janxdis.2016.11.008

qgraph: Epskamp, S., Cramer, A. O. J., Waldorp, L. J., Schmittmann, V. D., & Borsboom, D. (2012). qgraph: Network Visualizations of Relationships in Psychometric Data. Journal of Statistical Software, 048(i04). https://econpapers.repec.org/article/jssjstsof/v_3a048_3ai04.htm

bootnet: Epskamp, S., & Fried, E. I. (2018). A tutorial on regularized partial correlation networks. Psychological Methods, 23(4), 617–634. https://doi.org/10.1037/met0000167


